---
title: "Capstone Project"
author: "Gideon Ugboma"
date: "10/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Case Study: Cyclistic Bike-Share Analysis**

How does a bike-share company navigate speedy success?

The goal is to follow the steps of the data analysis process: ask, prepare, process, analyze, share, and act to answer the key business question: How do annual members and casual riders use Cyclistic bikes differently?

### **BACKGROUND**

**THE COMPANY**: Cyclistic, a fictional bike-share company based in Chicago, with a  fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago.
Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members.
Annual members are more profitable. Design marketing strategies aimed at converting casual riders into annual members. 

MY ROLE: A Junior data analyst working in the marketing analyst team.

### STAKEHOLDERS 
*	The director of marketing and my manager
*	Cyclistic marketing analytics team
*	Cyclistic executive team

## **ASK**

Business Task: Design marketing strategies aimed at converting casual riders into annual members.
The assignment is to find out: How do annual members and casual riders use Cyclistic bikes differently? And produce a report with the following deliverables:
1. A clear statement of the business task
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. A summary of your analysis
5. Supporting visualizations and key findings
6. Your top three recommendations based on your analysis

## **PREPARE**

Download the previous 12 months of Cyclistic trip historical data (from April, 2020 – March, 2021) from here [link](https://divvy-tripdata.s3.amazonaws.com/index.html).

Data consists of 12 separate cvs files with 13 variable.
The datasets have a different name because Cyclistic is a fictional company. The data has been made available by Motivate International Inc. You can find the license here [link](https://www.divvybikes.com/data-license-agreement) and the data is public.
Does your data ROCCC? : The data meets the ROCCC critaria.
*	Reliable - yes, not biased, 
*	Original - yes, the original data is in the public domain
*	Comprehensive - yes, not missing important information
*	Current - yes, it is updated monthly
*	Cited – yes

## **PROCESS**



### Setting up my environment
 Setting up my environment by installing and loading packages
 
 
```{r}
list.files(path = "./Bike-share Project/")
```
 

```{r installing packages}
library(skimr)
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
library(scales)
```

### Data Source Click Here [link](https://divvy-tripdata.s3.amazonaws.com/index.html)

```{r importing data}
apr_20 <- read.csv("./Bike-share Project/trip_data/202004-divvy-tripdata.csv")
may_20 <- read.csv("./Bike-share Project/trip_data/202005-divvy-tripdata.csv")
jun_20 <- read.csv("./Bike-share Project/trip_data/202006-divvy-tripdata.csv")
jul_20 <- read.csv("./Bike-share Project/trip_data/202007-divvy-tripdata.csv")
aug_20 <- read.csv("./Bike-share Project/trip_data/202008-divvy-tripdata.csv")
sep_20 <- read.csv("./Bike-share Project/trip_data/202009-divvy-tripdata.csv")
oct_20 <- read.csv("./Bike-share Project/trip_data/202010-divvy-tripdata.csv")
nov_20 <- read.csv("./Bike-share Project/trip_data/202011-divvy-tripdata.csv")
dec_20 <- read.csv("./Bike-share Project/trip_data/202012-divvy-tripdata.csv")
jan_21 <- read.csv("./Bike-share Project/trip_data/202101-divvy-tripdata.csv")
feb_21 <- read.csv("./Bike-share Project/trip_data/202102-divvy-tripdata.csv")
mar_21 <- read.csv("./Bike-share Project/trip_data/202103-divvy-tripdata.csv")
```


### Combine all Datasets into one
Here all individual datasets was merged into one

```{r merge all 12 datasets}
all_rides <- rbind(apr_20,may_20,jun_20,jul_20,aug_20,sep_20,oct_20,nov_20,dec_20,jan_21,feb_21,mar_21)
```

### Clean data to remove all empty cells and filter out all empty start station name

```{r}
all_rides <- janitor::remove_empty(all_rides,which = c("cols"))
all_rides <- janitor::remove_empty(all_rides,which = c("rows"))
all_rides <- all_rides %>% filter(start_station_name !="")
```


### Check Data Structure
Checking data structures to make sure the data formatted correctly

```{r str}
str(apr_20)
str(may_20)
str(jun_20)
str(jul_20)
str(aug_20)
str(sep_20)
str(oct_20)
str(sep_20)
str(oct_20)
str(nov_20)
str(dec_20)
str(jan_21)
str(feb_21)
str(mar_21)
```

### Covert Date/Time from character to Date/Time
Here I converted the date/time stamp from character to correct format 
Add columns that list the year, month, day of each ride

```{r}
all_rides$date <- as.Date(all_rides$started_at)
all_rides$started_at <- lubridate::ymd_hms(all_rides$started_at)
all_rides$ended_at <- lubridate::ymd_hms(all_rides$ended_at)
all_rides$start_hour <- lubridate::hour(all_rides$started_at)
all_rides$end_hour <- lubridate::hour(all_rides$ended_at)

all_rides$month <- format(as.Date(all_rides$date), "%B")
all_rides$day <- format(as.Date(all_rides$date), "%d")
all_rides$year <- format(as.Date(all_rides$date), "%y")
all_rides$day_of_week <- format(as.Date(all_rides$date), "%a")
```


### Calculating ride length and creating new columns


```{r}
all_rides$ride_length <- difftime(all_rides$ended_at,all_rides$started_at, units = "mins")
all_rides$Hours <- difftime(all_rides$ended_at,all_rides$started_at,units = c("hours"))
all_rides$Minutes <- difftime(all_rides$ended_at,all_rides$started_at,units = c("mins"))

```



```{r}
str(all_rides)

```


### Convert ride_length to numeric from factor


```{r}
is.factor(all_rides$ride_length)
all_rides$ride_length <- as.numeric(as.character(all_rides$ride_length))
is.numeric(all_rides$ride_length)

```

### Remove trips where the ride length is <= 0 or more than one day (24 * 60 = 1440 min)

```{r}
all_rides2 <- all_rides[!(all_rides$ride_length > 1440 | all_rides$ride_length <= 0),]
```


```{r}
str(all_rides2)

```


```{r}
View(all_rides)
```


```{r}
View(all_rides2)
```


### Type of membership
Here I determine the number of each membership types
 
```{r}
table(all_rides$member_casual)

```

### Combine start and end stations

```{r}
all_stations <- rbind(data.frame("stations" = all_rides2$start_station_name, 
                                     "member_casual" = all_rides2$member_casual),
                          data.frame("stations" = all_rides2$end_station_name,
                                     "member_casual" = all_rides2$member_casual))

```
### Removing entries with no station name

```{r}
all_stations2 <- all_stations[!(all_stations$stations == "" | is.na(all_stations$stations)),]
```


### Separate the data frame by rider type 

```{r}
all_stations_member <- all_stations2[all_stations2$member_casual == 'member',]
all_stations_casual <- all_stations2[all_stations2$member_casual == 'casual',]
```

## **ANALYZE**
Station analysis

### Top 10 popular stations all, members, and casual riders

```{r}
top_10_stations <- all_stations2 %>%
  group_by(stations) %>% 
  summarise(station_count = n()) %>% 
  arrange(desc(station_count)) %>% 
  slice(1:10)

top_10_station_member <- all_stations_member %>% 
  group_by(stations) %>% 
  summarise(station_count = n()) %>% 
  arrange(desc(station_count)) %>% 
  head(n=10)

top_10_station_casual <- all_stations_casual %>% 
  group_by(stations) %>% 
  summarise(station_count = n()) %>% 
  arrange(desc(station_count)) %>% 
  head(n=10)
```

### Top 20 start stations for casual riders

```{r}
all_rides2 %>% 
  group_by(start_station_name, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(start_station_name != "", member_casual != "member") %>% 
  arrange(-number_of_rides) %>% 
  head(n=20)
```


### Ride Length Analysis


```{r}
all_rides2 %>% group_by(member_casual) %>% 
  summarise(min_ride_length = min(ride_length), 
            max_ride_length = max(ride_length),
            mean_ride_length = mean(ride_length),
            median_ride_length = median(ride_length),
            mode_ride_length = mode(ride_length),
            count = n()
            ) %>% ungroup()
```


```{r}
summary(all_rides2$ride_length)

```


```{r}
all_rides2 %>% 
  summarise(min_ride_length = min(ride_length), 
            max_ride_length = max(ride_length),
            mean_ride_length = mean(ride_length),
            median_ride_length = median(ride_length),
            count = n())

```



### Compare the ride length between members and casual riders

```{r}
aggregate(all_rides2$ride_length ~ all_rides2$member_casual, FUN = mean)
aggregate(all_rides2$ride_length ~ all_rides2$member_casual, FUN = median)
aggregate(all_rides2$ride_length ~ all_rides2$member_casual, FUN = max)
aggregate(all_rides2$ride_length ~ all_rides2$member_casual, FUN = min)

```


### Average ride length by each day of the week for members vs. casual riders


```{r}

aggregate(all_rides2$ride_length ~ all_rides2$member_casual + all_rides2$day_of_week, FUN = mean)

```



### Arrange days of the week in order

```{r}
all_rides2$day_of_week <- ordered(all_rides2$day_of_week, 
                                  levels = c("Mon", "Tue", "Wed",
                                               "Thu", "Fri", "Sat", "Sun"))

aggregate(all_rides2$ride_length ~ all_rides2$member_casual + all_rides2$day_of_week, FUN = mean)

```


### Median ride length between members and casual riders for each day of week

```{r}
aggregate(all_rides2$ride_length ~ all_rides2$member_casual + all_rides2$day_of_week, FUN = median)
```

### Number of rides between members and casual riders for each day of week

```{r}
all_rides2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(day_of_week)
```
### Arrange months in order

```{r}
all_rides2$month <- ordered(all_rides2$month, 
                              levels = c("January", "February", "March",
                                         "April", "May", "June",
                                         "July", "August", "September",
                                         "Octobor", "November", "December"))
```


### Calculate the average ride length by month for members vs. casual riders

```{r}
aggregate(all_rides2$ride_length ~ all_rides2$member_casual + all_rides2$month, FUN = mean)
```


### Number of rides between members and casual riders for each month

```{r}
all_rides2 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(month) %>% ungroup()
```


### Compare bike type preference between casual riders and members
 
```{r}
all_rides2 %>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(number_of_rides = n()) %>% ungroup()
```


### Number of docked_bike rides by casual riders and members for each day of week
 
```{r}
all_rides2 %>% 
  filter(rideable_type == 'docked_bike') %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(day_of_week) %>% ungroup()
```
 
## Docked_bike rides by casual riders for each day of week

```{r}
all_rides2 %>% 
  filter(rideable_type == 'docked_bike', member_casual == 'casual') %>% 
  group_by(day_of_week) %>% 
  summarise(number_of_rides = n()) %>% ungroup()
```



```{r}
all_rides3 <- all_rides2 %>% group_by(weekly = floor_date(date,"week"),start_hour) %>% 
  summarise(
    Minutes = sum(Minutes),
    Mean = mean(Minutes),
    Median = median(Minutes),
    Max = max(Minutes),
    Min = min(Minutes),
    Count = n()
  ) %>% ungroup()
```


```{r}
bikestype <- all_rides2 %>% group_by(rideable_type,weekly = floor_date(date,"week")) %>% 
  summarise(
    Minutes = sum(Minutes),
    Mean = mean(Minutes),
    Median = median(Minutes),
    Max = max(Minutes),
    Min = min(Minutes),
    Count = n()
  ) %>% ungroup()
```


### Count by Bike Type(Total by Week)

```{r}
table(bikestype$rideable_type)

```


```{r}
xtabs(all_rides3$Count ~all_rides3$start_hour)
```


```{r}
View(all_rides3)
```

### Summary of Hourly Counts

```{r}
summary(all_rides3$Count)
```




```{r}
all_rides3$monthly <- lubridate::month(all_rides3$weekly)
```


```{r}
all_rides3$CountMA <- forecast::ma(all_rides3$Count,28)
```

## **SHARE**

### Creating Data Visualizations


```{r}
all_rides2 %>% count(start_station_name,sort = TRUE) %>% top_n(20) %>% ggplot() + 
  geom_col(aes(x=reorder(start_station_name,n),y=n,fill="peru")) + coord_flip() + scale_alpha_continuous(labels = comma) + labs(title = "Top 20 Start Station by Ride Count",
 X = "Count of Rides",y = "Station Name")                
```

### Top 10 Stations by Casual Riders

```{r}
ggplot(data = top_10_station_casual) +
  geom_col(aes(x = reorder(stations, station_count), y = station_count), fill = "purple") +
  labs(title = "Top 10 Stations by Casual Riders", x = "", y = "Number of Rides") + 
  scale_y_continuous(labels = scales::comma) +
  coord_flip()
```

### Top 10 Stations Used by Members

```{r}
ggplot(data = top_10_station_member) +
  geom_col(aes(x = reorder(stations, station_count), y = station_count,fill = "peru")) +
  labs(title = "Top 10 Stations Used by Members", y = "Number of Rides", x = "") +
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  theme_minimal() 
```

### Bike Use by Casual Riders vs. Members for each Day of Week across the Year

```{r}
all_rides2 %>% 
  group_by(month, day_of_week, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::comma) +
  facet_grid(member_casual~month) +
  labs(x = "Day of Week", y = "Number of Rides", fill = "Member/Casual",
       title = "Bike Use by Casual Riders vs. Members for each Day of Week across the Year", fill = 'Member/Casual') + theme(axis.text.x = element_text(angle = 90)) + theme()
```

### Count of ride per day

```{r}
all_rides3 %>% ggplot() + geom_col(aes(x = weekly,y = Count,fill="peru")) + scale_y_continuous(labels = comma) + 
  labs(title="Count of ride per day",y="Average rides per day")
```

### Count of ride by Hour

```{r}
all_rides3 %>% ggplot() + geom_col(aes(x = start_hour,y = Minutes,fill="peru")) + scale_y_continuous(labels = comma) + 
  labs(title="Count of ride by Hour",y="Rides per Hour")
```


### Count of ride by Bike Type

```{r}
ggplot(bikestype) + geom_area(aes(x=weekly,y=Count,fill=rideable_type)) + scale_y_continuous(labels=comma) + 
  labs(title="Count of ride by Bike Type")
  
```

### Total Ride Minutes by week

```{r}
ggplot(bikestype) + geom_col(aes(x = weekly,y = Minutes,fill = "peru")) + scale_y_continuous(labels = comma) +
  facet_wrap(~rideable_type) + labs(title = "Total Ride Minutes by week")
```

### Count of rides by Rider Type

```{r}
ggplot(bikestype) + geom_col(aes(x=weekly,y=Count,fill=rideable_type)) +  scale_y_continuous(labels=comma) + 
  labs(title="Count of rides by Rider Type")
  
```

### Ride Minutes by Bike Types and Week

```{r}
ggplot(bikestype ,aes(x=weekly, y=Minutes,fill=rideable_type)) + geom_area(stat ="identity",position=position_dodge(),alpha = 0.75) +
  scale_y_continuous(labels = comma) + labs(title="Ride Minutes by Bike Types and Week", y="Bike Trip in Minutes")
```



### **KEY FINDINGS**

*	There are more casual riders on weekends than week days.
*	During the summer months the number of casual riders steadily increase between the months of June to August
*	Docked bikes are the most popular and classic bikes are the least popular
*	Between 3 PM to 6 PM is the peak time with highest number of rides during the day
*	The mean ride length of casual riders is two times longer than members
*	The top 3 start stations for casual riders are 1. Streeter Dr. & Grand Ave. 2. Lake Shore Dr.& Monroe St. and 3.   
  Millennium Park

## **ACT**

### Recommendations:
*	Offer attractive discount on memberships during summer months and especially weekends to attract casual riders.
*	Marketing campaigns targeting top stations used by casual riders may increase the rate of subscription.
*	Form partnership with local businesses around the top 10 stations used by casual riders. 
*	Carryout marketing campaigns on digital media, cycling apps, billboards and local businesses on key cycling routes.

